Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-02-15 text/paper2/ms.tex
small samples text

The potential of the inner galaxy is likely well-described by a (set of) closely-integrable potentials, thus the dynamics of collisionless systems is plausibly captured by a transformation to some integral-of-motion-space, e.g., action-angle coordinates. However there is uncertainty in the exact shape, radial profile, and graininess of the outer halo. To test the extent to which the outer halo is well-fit by integrable potentials, we cannot use methods that \emph{assume} integrability. In this article, we present a fully probabilistic method for measuring properties of a potential that only relies on numerical orbit integration in ordinary phase-space. The model uses individual stars as constraints on the potential, and can therefore handle very small samples of stars. Generative models that instead rely on matching densities (in phase-space, action-space, or observed coordinates) must also include fine details about the progenitor (e.g., mass function or structure) and a model for the background of stars. Our method also naturally extends to using multiple debris structures \citep[see][who illustrate the power of using multiple streams to simultaneously constrain the potential]{deg14}.

[Sometimes stellar samples might get extremely small.  For example,
  \gaia\ might find some very low-mass cold stellar streams.  For
  another, there might only be a couple RRL stars in a particular
  stream, but those RRL stars might be individually extremely well
  measured in six-space.]

[Even if you have many stars, some might be orders of magnitude more constraining (Bovy, ..., Hogg)]

>>>>

The potential of the inner galaxy is likely well-described by a (set of) nearly-integrable potentials, thus the dynamics of collisionless systems is plausibly captured by a transformation to some integral-of-motion-space, e.g., action-angle coordinates. However there is uncertainty in the exact shape, radial profile, and graininess of the outer halo. To test the extent to which the outer halo is well-fit by integrable potentials, we cannot use methods that \emph{assume} integrability. In this article, we present a fully probabilistic method for measuring properties of a potential that only relies on numerical orbit integration in ordinary phase-space. The model uses individual stars as constraints on the potential, and can therefore handle very small samples of stars. Generative models that instead rely on matching densities (in phase-space, action-space, or observed coordinates) must also include fine details about the progenitor (e.g., mass function or structure) and a model for the background of stars. Our method also naturally extends to using multiple debris structures \citep[see][who illustrate the power of using multiple streams to simultaneously constrain the potential]{deg14}.

Although the \gaia\ project will deliver hundreds of millions of stars useful for dynamical inference,
  it still might be that the most relevant stellar samples for inferring the Milky Way potentail might be very small in the end.
One reason is that if distances turn out to be important, stars that produce good distance estimates might be much more
  valuable than typical stars in the sample.
At halo distances, \gaia's parallaxes are not expected to be better than the distance indicators produced by the best
  variable stars observed at the best wavelengths; for example, near-infrared observations of RR Lyrae stars
  can in principle deliver XX percent photometric distances at essentially all halo distances (CITATIONS).
These valuable stars are rare (and age and metallicity dependent);
  there could be many cold structures in the Milky Way halo that are highly constraining
  on the potential in principle, but which contain only a few good distance-indicating members.
That said, it is not yet known what the trade-offs are between having many stars at low precision and a few at high precision,
  nor is it known how valuable distance information really is, when a structure contains many precisely observed members.

Another reason to think about small samples is more conceptual:
In dynamical inferences that go beyond simple virial or Jeans arguments,
  inference depends on the non-linear dynamical evolution of populations in phase space.
The nonlinearities, combined with heteroskedastic data and requirements to marginalize over phase-space distribution functions,
  lead naturally to situations in which some stars are \emph{far more informative} than others.
For a simple example, in a toy study of the Solar System,
  phase-space distribution marginalization was shown to make some planets are far more informative than others
  about the gravitational potential (CITE Bovy et al).
This is expected to be a generic feature in dynamical problems;
  it might be that a very small fraction of \gaia's catalog entries
  deliver a large fraction of the information about the Milky Way's potential and formation history.
